---
title: "Project"
format: html
execute:
  echo: false
---

```{r}
#| label: setup
#| message: false

library(tidyverse)
```

## Most Popular Dog Breeds

::: {.callout-note}
This graph shows the ten most popular dog breeds in 2020 based on American Kennel Club (AKC) registration records. The **Breed** variable lists each dog breed, and the **2020 Rank** variable measures its popularity standing, where a rank of 1 represents the most popular breed (Labrador Retrievers).

These breeds lead in popularity because they combine exceptional **family affection** (creating loyal, loving companions) with high **trainability** (making them easier to housebreak and socialize). In addition, their manageable **energy levels** allow them to adapt well to typical family routines and living spaces.
:::

```{r}
#| label: most-popular-dogs
#| message: false

# Read the breed rankings dataset
ranks <- read_csv("data/breed_rank_all.csv")

# Select the top 10 breeds for 2020 and order by rank
top_dogs <- ranks |>
  slice_min(`2020 Rank`, n = 10) |>
  mutate(Breed = fct_reorder(Breed, `2020 Rank`, .desc = TRUE))

# Plot the top 10 dog breeds
ggplot(top_dogs, aes(x = `2020 Rank`, y = Breed)) +
  geom_point(size = 3.5, color = "steelblue") +
  scale_x_continuous(breaks = 1:10) +
  labs(
    title = "Top 10 Most Popular Dog Breeds (2020)",
    subtitle = "Rankings from the American Kennel Club",
    x = "AKC Rank (1 = Most Popular)",
    y = NULL
  ) +
  theme_minimal()
```

## Combining Rankings with Traits

```{r}
#| label: join-data
#| message: false

# Read the traits dataset
traits <- read_csv("data/breed_traits.csv")

# Clean whitespace in breed names and merge the datasets
combined_dogs <- ranks |>
  mutate(Breed = str_squish(Breed)) |>
  left_join(traits |> mutate(Breed = str_squish(Breed)), by = "Breed")

# Reshape key traits for the top 10 breeds
top_traits <- combined_dogs |>
  slice_min(`2020 Rank`, n = 10) |>
  select(Breed, `2020 Rank`, `Affectionate With Family`, `Trainability Level`, `Energy Level`) |>
  pivot_longer(
    cols = c(`Affectionate With Family`, `Trainability Level`, `Energy Level`),
    names_to = "Trait",
    values_to = "Score"
  ) |>
  mutate(Breed = fct_reorder(Breed, `2020 Rank`, .desc = TRUE))

# Plot the traits
ggplot(top_traits, aes(x = Score, y = Breed, fill = Trait)) +
  geom_col(position = "dodge") +
  scale_x_continuous(limits = c(0, 5), breaks = 1:5) +
  labs(
    title = "Key Traits of the Top 10 Most Popular Dog Breeds",
    subtitle = "Comparing affection, energy, and trainability (scores 1 to 5)",
    x = "Trait Score",
    y = NULL
  ) +
  theme_minimal() +
  theme(legend.position = "bottom")
```

### Why These Breeds Dominate

These three traits strongly explain why these breeds lead in popularity:

- **High Family Affection (average 4.7/5):** Almost every top breed scores a maximum 5 for affection. Pet owners prioritize companionship, loyalty, and gentleness with family members.
- **High Trainability (average 4.5/5):** Most top dogs score a 4 or 5 in trainability, making them easier to house-train, socialise, and handle in public settings compared to more independent breeds.
- **Manageable & Versatile Energy (average 3.8/5):** Top breeds offer a split that matches modern lifestyles: low-to-moderate energy breeds (like French Bulldogs and Bulldogs at 3/5) thrive in apartments, while high-energy breeds (like Labradors and German Shepherds at 5/5) appeal to active households.

